---
title: "Project 1: Checkpoint 4"
format: html
embed-resources: true
editor: source
code-tools: true
execute:
  warning: false
---

```{r}
#| label: packages-setup
#| message: false
#| warning: false
#| code-fold: true

library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(glue)
library(lubridate)
library(ggplot2)
library(knitr)
library(kableExtra)
library(scales)
```

## The Data

```{r}
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
```

```{r}
covid_cleaned_df <- read.csv("covid_cleaned.csv")
country_gdp <- read.csv("country_gdp.csv")
```

## Two Visualizations

**Measles vs. Covid-19 Cases in 2020**

```{r}
#| code-fold: true
#| fig-alt: "Horizontal bar graph comparing Covid-19 and measles cases in the 10 countries with the highest count of Covid-19 cases. Covid-19 cases significantly outnumber the measles cases."

measles_2020 <- cases_year |>
  filter(year == 2020) |>
  select(country, measles_total_2020 = measles_total)

covid_2020 <- covid_cleaned_df |>
  select(country, starts_with("X2020")) |>
  rowwise() |>
  mutate(covid_total_2020 = max(c_across(starts_with("X2020")), 
                                na.rm = TRUE)) |>
  ungroup() |>
  select(country, covid_total_2020)

comparison_2020 <- inner_join(measles_2020, covid_2020, by = "country") |>
  arrange(desc(covid_total_2020)) |>
  slice_head(n = 10) |>
  pivot_longer(cols = c(measles_total_2020, covid_total_2020), 
               names_to = "disease", values_to = "cases") |>
  mutate(disease = case_when(disease == "measles_total_2020" ~ "Measles", 
                             disease == "covid_total_2020"   ~ "COVID-19"), 
         country = fct_reorder(country, cases, .fun = max))

ggplot(comparison_2020, aes(y = country, x = cases, fill = disease)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  scale_x_log10(labels = label_number(scale_cut = cut_short_scale()), 
                breaks = 10^(0:8)) +
  scale_fill_manual(values = c("COVID-19" = "#D85A30", 
                               "Measles" = "#1D9E75")) +
  labs(title    = "Measles vs. COVID-19 Cases in 2020", 
       subtitle = "Top 10 countries by COVID-19 case count — log scale",
       x = "Total Cases (log scale)", y = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.text.y = element_text(size = 10))
```

**Concentration of Measles Cases Across Countries**

```{r}
#| code-fold: true
#| fig-alt: "Lollipop plot showing the total number of measles cases in all countries by year.  Most years have similar totals, however there is a significant peak in 2019."

global_totals <- cases_month |> 
  group_by(year) |> 
  summarize(total_cases = sum(measles_total, na.rm = TRUE))

ggplot(global_totals, aes(x = year, y = total_cases)) + 
  geom_segment(aes(yend = 0, 
                   xend = year), 
               color = "purple2") + 
  geom_point(size = 4, color = "purple4") + 
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  theme_minimal() + 
  labs(x = "Year", y = "Total Cases", 
       title = "Measles Cases Across the World Over Time", 
       subtitle = "Total Count of Global Cases by Year")
```

This lollipop plot visualizes the spread of total measles cases across the world by year. The peak in this graph is occurring at 2019. In the final report, we will explore further why this peak is happening and if there is a specific country that is responsible.

## Statistical Model

**RQ: Is there a significant difference in mean GDP growth between countries with high vs. low measles burden in 2019?**

To answer this research question, the model below fits a linear regression model between GDP growth and the total number of measles and rubella cases.

```{r}
cases_gdp_join <- inner_join(cases_year, country_gdp, 
                             by = join_by(year, iso3 == country_code)) |>
    mutate(
      high_measles = measles_lab_confirmed > median(measles_lab_confirmed)
      ) |>
  filter(year == 2019)
```

```{r}
t.test(gdp_growth ~ high_measles, data = cases_gdp_join)
```

Conclusion: Using the 2019 data, a year marked by a notable rise in measles cases, countries were classified as high measles burden if their lab-confirmed measles cases exceeded the median of 48 cases. A two sample t-test resulted in no significant difference in mean GDP growth between countries with high and low measles burden. Due to a t-value = -1.04 and a p-value = 0.299 > alpha = 0.05, we fail to reject the null hypothesis.
















